---
title: "Dead wood stocks and fluxes for Barro Colorado Island 50-ha forest dynamic plot, 2017-2024"
author: "Lily Pitcher and Helene C. Muller-Landau"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
                       #"/Users/lilyp/OneDrive - Smithsonian Institution/Documents/D-woody debris/FILES FOR CODE")
```




```{r, include = FALSE}
library(ggplot2)
library(cowplot)
library(knitr)
library(kableExtra)
library(report)
library(tidyr)
library(RColorBrewer)
library(dplyr)
library(lubridate)
library(zoo)
library(here)
library(Rmisc)

rm(list=ls())
```

### Citation and License

This report is part of the following data and code publication, which should be cited as follows:

It is licensed under CC BY 4.0.


### Contributions and Acknowledgments

Lily Pitcher conducted data cleaning, performed the analyses and wrote this Rmarkdown report.  Helene C. Muller-Landau co-designed the study, supervised the field work,  supervised the analyses, and edited the report.  Pablo Ramos and Paulino Villareal collected the field data.  Markku Larjavaara co-designed the study.  This work builds on previous data cleaning and analyses by Evan Gora (see in Gora et al. 2019).  This study was funded by Smithsonian ForestGEO and the HSBC Climate Partnership. 


### Study site and sample design
 
 This document reports the estimated above ground Coarse deadwood stocks of dead trees in the 50ha plot of Barrow Colorado Island, STRI, Panama. Woody debris is estimated for 100 (40X40m) within the 50ha through annual field studies which comprised of the following studies:
 
Coarse Woody Debris (CWD) dynamics:
CWD is defined as deadwood with a diameter equal or greater than 200cm. Four 40-m transects, 2 in each of 2 perpendicular directions, span each 40x40 m subplot. For fallen deadwood pieces the diameter is measured at the point where the piece crosses the transect.In addition, each piece is marked to allow it to be tracked over time and distinguished from newly fallen debris in the next census. We also record the tag number of the tree from which the piece came, if discernible. We also categorize into branch or trunk.
We inventory standing dead trees within the 40x40 sub-plots.For each standing dead tree greater than 200 mm in diameter, we measure
dbh and height. In addition, we categorize the abundance of branches
remaining (<10 %, 10 – 90 % or > 90 %). In the case of stumps, we measure diameter at half way between the top and ground. We try to record the tag.

 Further details of sampling design and protocols can be found here:
 
https://forestgeo.si.edu/protocols/woody-debris



```{r, include=FALSE}
#load in subplot coordinates and habitat classes to plot the 50ha
fnhabitat<-"Data0_Raw/bci_q20habitat_classes"
fnsubplot<-"Data0_Raw/subplot_codes.txt"
subplot<-read.table(fnsubplot, header=TRUE, sep="\t")
habitat<-read.table(fnhabitat,header=TRUE, sep="\t")
subplot_sep<-separate(subplot, col = subplot_code, into = c("x", "y"), sep = ",")
subplot_sep$x<-as.numeric(subplot_sep$x)
subplot_sep$y<-as.numeric(subplot_sep$y)
habitat$x<-habitat$x/20
habitat$y<-habitat$y/20
```

```{r, echo=FALSE, warning=FALSE, fig.width=9.5}
#plotting the 100 subplots across the 50 ha plot
colours<-c("#228B22", "#D2B48C","#66CDAA","#2E8B57", "#64A8D1", "#556B2F", "#ADFF2F")
ggplot() +
    geom_tile(data = habitat,aes(x=x, y=y, fill=habitat, group=habitat)) +
  geom_tile(data=subplot_sep, aes(x=x, y=y), colour = "red", fill=NA, size=0.5)+
  scale_fill_manual(
    values = colours) +
  labs(title="Map displaying the dynamic subplot locations and habitat zones across the 50ha plot")+
  theme_bw() +
  coord_fixed(ratio=1) + # ensures 1:1 aspect ratio
  theme(
    legend.title = element_text(size = 7),
    legend.key.size = unit(1, 'cm'),
    axis.text = element_blank(),
    axis.title = element_blank(),# Removelabels
    strip.text = element_text(size = 10), # Adjust facet label size if needed
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )
rm(subplot_sep)
```
This map shows the 100 dynamic subplots across the 50ha plot, represented by the red squares. Colours represent the different habitat zones across the 50 ha plot.


 
```{r, include = FALSE}
## Load fallen and standing CWD data
FNFallen_DATA <-"Data3_Processed/processed_CWD40_fallen_17to24.csv"
FNStanding_DATA <-"Data3_Processed/processed_CWD40_standing_17to24.csv"

fallen_data<-read.csv(FNFallen_DATA,header=TRUE) 
names(fallen_data)[names(fallen_data)=="year"] <- "yearcol"

standing_data<-read.csv(FNStanding_DATA,header=TRUE)
names(standing_data)[names(standing_data)=="year"] <- "yearcol"
```


```{r, include=FALSE}
# 1. FALLEN COARSE WOODY DEBRIS — MASS AND VOLUME CALCULATIONS

# Calculations to estimate the mass and volume of fallen coarse woody debris per subplot per year
# Output units:
# - Mass: Mg/ha
# - Volume: m³/ha

SUBPLOTCALCULATIONSFALLEN<-function(indata){
indata<-indata %>%
  dplyr::group_by(yearcol, subplot_code) %>% 
  # 160 is the total length of the transects in one subplot (m) 
  dplyr::summarise(
    mass.kg.m2=((pi/(2*160))*sum(pen_crossmass.kgm)),           
    vol.m3.m2=((pi^2/(8*160))*sum(diam_sqr_m))) %>%
  
  dplyr::mutate(
    mass.Mg.ha=mass.kg.m2*10, 
    vol.m3.ha=vol.m3.m2*10000)%>%
  
  dplyr::select(-c(mass.kg.m2,vol.m3.m2))

return(indata)
}
#Apply function to fallen data
subplot_fallen<-SUBPLOTCALCULATIONSFALLEN(fallen_data)


# Add zeros for subplot-year combinations with no fallen CWD recorded
add_zeros<- function(indata, subplot, outdata) {
  all_subplots<- unique(subplot$subplot_code)
  all_years<-unique(indata$yearcol)
  all_combinations<- expand.grid(subplot_code = all_subplots, yearcol = all_years)
      #b.merge
  outdata <- merge(all_combinations, indata, by = c("subplot_code", "yearcol"), all.x = TRUE)
  outdata<- outdata %>%
      dplyr::mutate(across(where(is.numeric), ~replace(., is.na(.), 0)))
  return(outdata)
}
subplot_fallen<-add_zeros(subplot_fallen, subplot, subplot_fallen)

# Calculate mean and standard deviation of mass and volume per year
year_fallen<-subplot_fallen %>%
  dplyr::group_by(yearcol) %>%
  dplyr::summarise(mean_mass.Mg.ha=mean(mass.Mg.ha),
                   sd_mass.Mg.ha=sd(mass.Mg.ha), 
                   mean_vol.m3.ha = mean(vol.m3.ha),
                   sd_vol.m3.ha= sd(vol.m3.ha) )
```


```{r, include=FALSE}
# Identify new vs. existing pieces based on year of first observation
# Used to distinguish inputs (recruitment) from residual (remaining) material

flux_calculation<-function(indata){
  yearly_count <-indata %>%
dplyr::group_by(yearcol) %>%
dplyr::summarise(count_per_year = n())
  
# Determine year of first appearance for each unique piece
new_codes_by_year <- indata %>%
  arrange(yearcol) %>%
dplyr::group_by(code_of_piece) %>%
dplyr::mutate(first_appearance = min(yearcol)) %>%  
  ungroup() 

# Summarise how many new pieces appear per year
summary_new_codes<-new_codes_by_year%>%
 dplyr::group_by(yearcol)%>%
 dplyr::summarise(new_codes = sum(first_appearance == yearcol))

stem_count <- yearly_count %>%
  left_join(summary_new_codes, by = "yearcol")

# Separate new and existing pieces
old_data<- new_codes_by_year %>%
  filter(first_appearance < yearcol)
new_data<- new_codes_by_year %>%
  filter(first_appearance == yearcol)

return(list(stem_count=stem_count, old_data=old_data, new_data=new_data))
}

```

```{r, include=FALSE}
# Separate fallen wood into existing (old) and new (input) components
combined_fallen<-flux_calculation(fallen_data)
old_raw_fallen<-combined_fallen$old_data
new_raw_fallen<-combined_fallen$new_data
stem_count_fallen<-combined_fallen$stem_count

# Calculate subplot-level mass and volume for old and new pieces
subplot_fallen_old<-SUBPLOTCALCULATIONSFALLEN(old_raw_fallen) 
subplot_fallen_new<-SUBPLOTCALCULATIONSFALLEN(new_raw_fallen)#

subplot_fallen_old<-add_zeros(subplot_fallen, subplot,subplot_fallen_old)
subplot_fallen_new<-add_zeros(subplot_fallen, subplot,subplot_fallen_new)


# Compute annual means for old (remaining) and new (input) fallen CWD
year_fallen_old<-subplot_fallen_old %>%
  dplyr::group_by(yearcol) %>%
  dplyr::summarise(remaining_mean_mass.Mg.ha=mean(mass.Mg.ha),
                   remaining_sd_mass.Mg.ha=sd(mass.Mg.ha),
                   remaining_mean_vol.m3.ha = mean(vol.m3.ha), 
                   remaining_sd_vol.m3.ha= sd(vol.m3.ha))

year_fallen_new<-subplot_fallen_new %>%
  dplyr::group_by(yearcol) %>%
  dplyr::summarise(input_mean_mass.Mg.ha=mean(mass.Mg.ha),
                   input_sd_mass.Mg.ha=sd(mass.Mg.ha),
                   input_mean_vol.m3.ha = mean(vol.m3.ha),
                   input_sd_vol.m3.ha= sd(vol.m3.ha))


rm(old_raw_fallen,new_raw_fallen, subplot_fallen_old )
```





```{r, include=FALSE}
 # 2. STANDING WOODY DEBRIS — VOLUME AND MASS CALCULATIONS

# Calculate subplot-level standing biomass and volume per m²,
# then scale to Mg/ha and m³/ha using subplot area (1600 m²)

SUBPLOTCALCULATIONSTANDING<-function(indata) {
indata<-indata %>%
  dplyr::group_by(yearcol, subplot_code) %>%
  dplyr::summarise(
    mass.kg.m2=sum(AGB.kg.dry.mass)/1600, 
    vol.m3.m2=sum(volume_m3)/1600)%>%
  
  dplyr::mutate(
    mass.Mg.ha=mass.kg.m2*10,
    vol.m3.ha=vol.m3.m2*10000)%>%
  
  dplyr::select(-c(mass.kg.m2,vol.m3.m2))
return(indata)}

# Apply functions to standing data
subplot_standing<-SUBPLOTCALCULATIONSTANDING(standing_data)
subplot_standing<-add_zeros(subplot_standing, subplot,subplot_standing)


# Calculate annual means and standard deviations
year_standing<-subplot_standing %>%
  dplyr::group_by(yearcol) %>%
dplyr::summarise(mean_mass.Mg.ha=mean(mass.Mg.ha),
                 sd_mass.Mg.ha=sd(mass.Mg.ha), 
                 mean_vol.m3.ha = mean(vol.m3.ha),
                 sd_vol.m3.ha= sd(vol.m3.ha) )

# Separate standing biomass into new recruits and existing individuals
combined_standing<-flux_calculation(standing_data)
old_raw_standing<-combined_standing$old_data
new_raw_standing<-combined_standing$new_data
stem_count_standing<-combined_standing$stem_count

# Subplot-level calculations for standing old vs. new components
subplot_standing_old<-SUBPLOTCALCULATIONSTANDING(old_raw_standing)
subplot_standing_new<-SUBPLOTCALCULATIONSTANDING(new_raw_standing)
subplot_standing_old<-add_zeros(subplot_standing, subplot,subplot_standing_old)
subplot_standing_new<-add_zeros(subplot_standing, subplot,subplot_standing_new)

# Annual averages of old and new standing components
year_standing_old<-subplot_standing_old %>%
 dplyr::group_by(yearcol) %>%
 dplyr::summarise(remaining_mean_mass.Mg.ha=mean(mass.Mg.ha),
                  remaining_sd_mass.Mg.ha=sd(mass.Mg.ha),
                  remaining_mean_vol.m3.ha = mean(vol.m3.ha),
                  remaining_sd_vol.m3.ha= sd(vol.m3.ha))

year_standing_new<-subplot_standing_new %>%
 dplyr::group_by(yearcol) %>%
dplyr::summarise(input_mean_mass.Mg.ha=mean(mass.Mg.ha),
                 input_sd_mass.Mg.ha=sd(mass.Mg.ha),
                 input_mean_vol.m3.ha = mean(vol.m3.ha),
                 input_sd_vol.m3.ha= sd(vol.m3.ha))

rm(old_raw_standing,new_raw_standing,  subplot_standing_old )
```
	




```{r, include=FALSE}
# Code to extract sampling dates and people involved in field data collection
# This section extracts key information used to summarise sampling effort during each annual CWD census.

# 1. Extract start and end dates from field records
date_extraction<-function(indata){
indata<-indata %>%
select(month, day, yearcol)
indata$date<-paste(indata$day,"_",indata$month,"_",indata$yearcol)
indata$date<-dmy(indata$date)
return(indata) }

# Extract sampling dates from fallen and standing datasets
collection_summary_fallen<-date_extraction(fallen_data)
collection_summary_fallen$census<-paste("fallen")
collection_summary_standing<-date_extraction(standing_data)
collection_summary_standing$census<-paste("standing")

# Combine all available sampling dates
date_summary<-merge(collection_summary_fallen, collection_summary_standing, by=c("month", "day", "yearcol", "date", "census"), all = TRUE)

# Summarise date ranges (start and end) of field sampling for each year
sample_date<-function(indata){
indata<-indata %>%
  dplyr::group_by(yearcol)%>%
  dplyr::summarise(start_date=min(date, na.rm=TRUE), end_date=max(date, na.rm=TRUE), .groups="drop")
indata$start_date<-  format(ymd(indata$start_date), "%d %B")
indata$end_date<-  format(ymd(indata$end_date), "%d %B")
return(indata)}
date_summary<-sample_date(date_summary)


# 2. Extract names of people involved in data collection from observer records
observer_filter<-function(indata){
  indata<-indata %>%
dplyr::select(yearcol, observer)
return(indata)
}

observer_fallen<-observer_filter(fallen_data)
observer_standing<-observer_filter(standing_data)
observer_data<-merge(observer_fallen, observer_standing, by=c("yearcol", "observer"),all=TRUE)

# Combine observer records from both censuses
observer_extraction<-function(indata){
   indata<-indata %>%
dplyr::group_by(yearcol) %>%
dplyr::summarise(observer = if_else(
      all(is.na(observer)),  "not recorded",
      paste(unique(na.omit(observer)), collapse = ", ")),
      .groups="keep")
return(indata)}
observer_summary<-observer_extraction(observer_data)

# Merge all summary data: dates, stem counts, and observers
collection_summary<-merge(date_summary, stem_count_fallen, by="yearcol")
collection_summary<-merge(collection_summary, stem_count_standing, by="yearcol")
collection_summary<-merge(collection_summary, observer_summary, by="yearcol")

rm(collection_summary_fallen, collection_summary_standing, date_summary, stem_count_fallen, stem_count_standing, observer_data, observer_summary, observer_fallen, observer_standing)

```



```{r, echo=FALSE}
# Display summary table of annual sampling efforts
kable(collection_summary, format = "html", row.names = FALSE,
      col.names = c("Year",
                    "Start Date",
                    "End Date",
                    "Total Fallen Pieces",
                    "New Fallen Pieces",
                    "Total Standing Pieces",
                    "New Standing Pieces",
                    "Field Team"),
  escape = FALSE) %>%  
kable_styling(bootstrap_options = c("striped", "hover", "bordered"),
                full_width = F,
                font_size = 15)%>%
                add_header_above(c("Summary of Annual CWD census" = 8))

rm(collection_summary)
```


**Table1**


Summary of census effort for each year including census period (start and end date), total number of pieces of wood measured each year for Fallen and Standing (Total Pieces) and the number of new pieces of dead wood (New pieces). Initials represent the following people involved in the sample collection: R:Pablo Ramos, V:Paulino Villarreal and E is both working as a team.

### Calculations of CWD stores and fluxes

This report calculates annual stocks and fluxes of standing and fallen coarse woody debris. The data were formatted, cleaned and processed prior to loading loading into this report using the code **clean0_bci50deadwood_rawtoformatted.RMD** and **clean1_bci50deadwood_formattedtocorrected.Rmd**. Calculations for individual sample pieceswere made using code in **density_calculation_bci50deadwood_correctedtoprocessed.Rmd** which also contains explanations of all calculations.  More information can be found in **calcexplanation_bci50deadwood_stocks&fluxes.Rmd**.  

LILY: update the file name above.  
Problem in table -input and total values are the same.



```{r, include=FALSE}
# Merge annual data for fallen woody debris (total, existing, and input estimates)
year_fallen<-merge(year_fallen, year_fallen_old, by="yearcol")
year_fallen<-merge(year_fallen, year_fallen_new, by="yearcol")

# Merge annual data for standing woody debris
year_standing<-merge(year_standing, year_standing_old, by="yearcol")
year_standing<-merge(year_standing, year_standing_new, by="yearcol")

# Swap values for 2017
# In the first census year (2017), what was labelled as "input" is actually the pre-existing samples
# Inputs cannot be separated from existing material for this initial year
swap_values<-function(indata) {
  row_to_swap<-indata$yearcol==2017
  
temp <- indata$input_mean_vol.m3.ha[row_to_swap]
indata$input_mean_vol.m3.ha[row_to_swap] <- indata$remaining_mean_vol.m3.ha[row_to_swap]
indata$remaining_mean_vol.m3.ha[row_to_swap] <- temp

temp <- indata$input_sd_vol.m3.ha[row_to_swap]
indata$input_sd_vol.m3.ha[row_to_swap] <- indata$remaining_sd_vol.m3.ha[row_to_swap]
indata$remaining_sd_vol.m3.ha[row_to_swap] <- temp

temp <- indata$input_mean_mass.Mg.ha[row_to_swap]
indata$input_mean_mass.Mg.ha[row_to_swap] <- indata$remaining_mean_mass.Mg.ha[row_to_swap]
indata$remaining_mean_mass.Mg.ha[row_to_swap] <- temp

temp <- indata$input_sd_mass.Mg.ha[row_to_swap]
indata$input_sd_mass.Mg.ha[row_to_swap] <- indata$remaining_sd_mass.Mg.ha[row_to_swap]
indata$remaining_sd_mass.Mg.ha[row_to_swap] <- temp

outdata<-indata
return(outdata)
}

```



```{r, include=FALSE}
#Creating .txt output files summarising CWD stocks and fluxes at subplot and whole-plot levels

# Rename input columns for clarity
RENAME<-function(indata){
indata<-indata %>%
  dplyr::rename(input.mass.Mgha=mass.Mg.ha,
                input.vol.m3ha=vol.m3.ha)
return(indata)}
# Apply to new fallen and standing inputs
subplot_fallen_new<-RENAME(subplot_fallen_new)
subplot_standing_new<-RENAME(subplot_standing_new)

# Merge total and input data at subplot level
subplot_fallen <-left_join(subplot_fallen, subplot_fallen_new, by = c("subplot_code", "yearcol"))
subplot_standing <-left_join(subplot_standing, subplot_standing_new, by = c("subplot_code", "yearcol"))

# Add indicator column to distinguish fallen and standing debris
subplot_fallen<-subplot_fallen %>%
  dplyr::mutate(type="fallen")
subplot_standing<-subplot_standing %>%
  dplyr::mutate(type="standing")

# Combine subplot-level data and export as .txt
subplot_all<-rbind(subplot_fallen,subplot_standing)

write.table(subplot_all, file = "Output/bci_CWD40_subplot_estimation_17to24.txt", sep = "\t",  col.names = TRUE, row.names = FALSE)


# Apply swap for 2017 and prepare whole-plot level estimates
year_fallen<-swap_values(year_fallen)
year_standing<-swap_values(year_standing)

#create new column that distinguishes fallen from standing woody debris
year_fallen2<-year_fallen %>%
  dplyr::mutate(type="fallen") %>%
  dplyr::select(yearcol,type, mean_mass.Mg.ha,sd_mass.Mg.ha,mean_vol.m3.ha,sd_vol.m3.ha,input_mean_mass.Mg.ha,input_sd_mass.Mg.ha,input_mean_vol.m3.ha,input_sd_vol.m3.ha)

year_standing2<-year_standing %>%
  dplyr::mutate(type="standing") %>%
  dplyr::select(yearcol,type, mean_mass.Mg.ha,sd_mass.Mg.ha,mean_vol.m3.ha,sd_vol.m3.ha,input_mean_mass.Mg.ha,input_sd_mass.Mg.ha,input_mean_vol.m3.ha,input_sd_vol.m3.ha)

# Combine and export whole-plot data as .txt
year_all<-rbind(year_fallen2,year_standing2 )
year_all<-year_all %>%
  arrange(yearcol)

write.table(year_all, file = "Output/bci_CWD40_annual_estimation_17to24.txt", sep = "\t",  col.names = TRUE, row.names = FALSE)

#rm(year_fallen2, year_standing2)
```







### Total estimates for CWD stocks, fluxes and residence time for the BCI 50-ha plot

```{r, include=FALSE}
# Compute mean annual total stock and input (volume and mass), and calculate residence time
# residence time calculated as total mass divided by input mass
ALL_SUMMARY<-function(indata){
indata<-indata %>%
  dplyr::select(mean_mass.Mg.ha, input_mean_mass.Mg.ha, mean_vol.m3.ha, input_mean_vol.m3.ha )%>%
  dplyr::summarise(mean_mass = mean(mean_mass.Mg.ha),
                   mass_input = mean(input_mean_mass.Mg.ha, na.rm=TRUE),
                   mean_vol = mean(mean_vol.m3.ha), 
                   vol_input = mean(input_mean_vol.m3.ha, na.rm=TRUE)) %>%
  dplyr::mutate(mean_residence_time= mean_mass/mass_input) 
return(indata)
}

#Applying functions to data
year_fallen_all<-ALL_SUMMARY(year_fallen2)
year_standing_all<-ALL_SUMMARY(year_standing2)
total_means<-rbind(year_fallen_all, year_standing_all)

rm(year_vol_fallen, year_mass_fallen, year_vol_standing, year_mass_standing,year_fallen_old, year_standing_old, year_fallen_new, year_fallen_old, total_means_fallen, total_means_standing)
```







```{r, echo=FALSE}
# Display mean summary statistics in a table
kable(total_means, format = "html", row.names=FALSE, digits=2,
      col.names = c("Mean Total Mass (Mg ha<sup>-1</sup>) ",
                    "Mean Input Mass (Mg ha<sup>-1</sup>y<sup>-1</sup>)",
                    "Mean Total Volume (m<sup>3</sup> ha<sup>-1</sup>)" ,
                    "Mean Input Volume (m<sup>3</sup> ha<sup>-1</sup> y<sup>-1</sup>)",
                    "Residence time 
                    (y<sup>-1</sup>)"),
  escape = FALSE) %>%  
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"),
                full_width = F,
                font_size = 15)%>%
                pack_rows("Fallen", 1, 1) %>%  # Group rows 1-8 as 'Fallen'
                pack_rows("Standing", 2, 2) %>%
                add_header_above(c("Coarse Woody Debris Summary Statistics" = 5))
```

**Table 2**

Estimated mean stocks and fluxes of fallen and standing CWD calculated by first averaging over the one hundred 40 x 40 m dynamics plots, then averaging over years (2017-2024). Estimated total stocks of deadwood (Mean total), and estimated annual deadwood input for both volume and mass per ha. Residence time represents the average residence time of dead wood in the plot which is calculated as mean stocks divided by the mean inputs.


### Annual variation in woody debris stocks and fluxes

```{r, include=FALSE}
# Combine standing and fallen coarse woody debris data
subplot_fallen<-subplot_fallen %>%
  mutate(Type=paste("Fallen")) 
subplot_standing<-subplot_standing %>%
  mutate(Type=paste("Standing"))

combined_total_data<-rbind(subplot_fallen, subplot_standing)
```



```{r, echo=FALSE, results='hide', message=FALSE}
library(Rmisc)
# Calculate mean and standard error of mass by year and type
summary_mean_sublot_mass<-summarySE(combined_total_data, measurevar = "mass.Mg.ha", groupvars = c("yearcol","Type"))

# Plot mass over time with confidence intervals
ggplot(summary_mean_sublot_mass, aes (x=factor(yearcol), y=mass.Mg.ha, fill=Type, colour = Type))+
geom_errorbar(aes(ymin=(mass.Mg.ha-1.96*se), ymax=(mass.Mg.ha+1.96*se)), width=.1, position = position_dodge(width = 0.6))+
geom_point(size=3, shape=21, colour="black", position = position_dodge(width = 0.6))+
  labs(title = "Mean (with 95% CI) Total Mass of Coarse Woody Debris by Year", 
       x = "Year", 
       y = expression(paste("Mean Mass (Mgha"^{-1},")"))) +  
  theme_bw() +  
  scale_fill_manual(values=c("#009E73", "#882255"))+
scale_colour_manual(values=c("#009E73", "#882255"))+
  scale_y_continuous(breaks = seq(0,20 , by = 2))+
  theme(
    axis.text.x = element_text()  
  )
```


**Figure 1** 


Annual variation in mean total mass of fallen and standing coarse woody debris (CWD) per hectare, with 95% confidence intervals (CIs calculated as 1.96 times the standard error of the mean over the one hundred 40 x 40 m subplots).




```{r, echo=FALSE, results='hide', message=FALSE}
# Calculate mean and standard error of volume by year and type
summary_mean_sublot_vol<-summarySE(combined_total_data, measurevar = "vol.m3.ha", groupvars = c("yearcol","Type"))

# Plot volume over time with confidence intervals
ggplot(summary_mean_sublot_vol, aes (x=factor(yearcol), y=vol.m3.ha, fill=Type, colour = Type))+
geom_errorbar(aes(ymin=(vol.m3.ha-1.96*se), ymax=(vol.m3.ha+1.96*se)), width=.1, position = position_dodge(width = 0.6))+
geom_point(size=3, shape=21, colour="black", position = position_dodge(width = 0.6))+
  labs(title = "Mean (with 95% CI) Total Volume of Coarse Woody Debris by Year", 
       x = "Year", 
       y = expression(paste("Mean Volume (m"^{3},"ha"^{-1},")"))) +  
  theme_bw() +  
  scale_fill_manual(values=c("#009E73", "#882255"))+
scale_colour_manual(values=c("#009E73", "#882255"))+
 scale_y_continuous(breaks = seq(0, 80, by = 5))+
  theme(
    axis.text.x = element_text()  
  )
```

**Figure 2** 


Annual variation in mean total volume of fallen and standing CWD per hectare, with 95% confidence intervals (CIs calculated as 1.96 times the standard error of the mean over the one hundred 40 x 40 m subplots).



### Distribution of CWD stocks and fluxes across the dynamic subplots
```{r, echo=FALSE,results='hide', message=FALSE, warning=FALSE}
# Plotting histogram of total mass of fallen CWD per subplot, coloured by year
# X-axis is log₂-transformed mass
ggplot(subplot_fallen, aes(x = log(mass.Mg.ha , base = 2), group=yearcol, fill = as.factor(yearcol)))+
  geom_histogram(colour = "black", 
                 bins = 40, position="stack")+
  labs(title = "Distribution of estimated Fallen CWD", 
       fill="yearcol",
       y = "Count", 
       x = expression(paste("Log2 Mass (Mgha"^{-1},")"))) +  
  theme_bw() +  
  scale_y_continuous(breaks = seq(0, 50, by = 5))+
  scale_x_continuous(breaks = seq(-3, 10, by = 1))
  
  
```


**Figure 3**


Distribution of log₂-transformed total estimated mass of falle coarse woody debris (CWD) per subplot, coloured by year. The x-axis represents log₂-transformed mass (Mg ha⁻¹), and the y-axis shows the count of subplots within each bin.

```{r, echo=FALSE,results='hide', message=FALSE, warning=FALSE}
# Plotting histogram of fallen CWD input mass per subplot, coloured by year (excluding 2017)
# X-axis is log₂-transformed mass
ggplot(subplot_fallen_new %>% filter(!yearcol=="2017"),aes(x = log(input.mass.Mgha , base = 2), group=yearcol, fill = as.factor(yearcol)))+
  geom_histogram(colour = "black", 
                 bins = 40, position="stack")+
 # facet_wrap("yearcol")+
  labs(title = "Distribution of estimated Input Mass of Fallen CWD", 
       fill="yearcol",
       y = "Count", 
       x = expression(paste("Log2 Mass (Mgha"^{-1},")"))) +  
  theme_bw() +  
  scale_y_continuous(breaks = seq(0, 20, by = 2))+
  scale_x_continuous(breaks = seq(-3, 10, by = 1))
  
  
```



**Figure 4**

Distribution of log₂-transformed input mass of fallen CWD per subplot (excluding 2017), coloured by year. The x-axis represents log₂-transformed mass (Mg ha⁻¹), and the y-axis shows the count of subplots within each bin.

```{r, echo=FALSE,results='hide', message=FALSE, warning=FALSE}
# Plotting histogram of total standing CWD mass per subplot, coloured by year
# X-axis is log₂-transformed mass
ggplot(subplot_standing, aes(x = log(mass.Mg.ha , base = 2), group=yearcol, fill = as.factor(yearcol)))+
  geom_histogram(colour = "black", 
                 bins = 40, position="stack")+
 # facet_wrap("yearcol")+
  labs(title = "Distribution of estimated Standing CWD", 
       fill="yearcol",
       y = "Count", 
       x = expression(paste("Log2 Mass (Mgha"^{-1},")"))) +  
  theme_bw() +  
  scale_y_continuous(breaks = seq(0, 50, by = 5))+
  scale_x_continuous(breaks = seq(-6, 10, by = 1))
  
  
```



**Figure 5**


Distribution of log₂-transformed total estimated mass of standing coarse woody debris (CWD) per subplot, coloured by year. The x-axis represents log₂-transformed mass (Mg ha⁻¹), and the y-axis shows the count of subplots within each bin.





```{r, echo=FALSE,results='hide', message=FALSE, warning=FALSE}
# Plotting histogram of standing CWD input mass per subplot, coloured by year 
# X-axis is log₂-transformed mass
ggplot(subplot_standing_new %>% filter(!yearcol=="2009"), aes(x = log(input.mass.Mgha , base = 2), group=yearcol, fill = as.factor(yearcol)))+
  geom_histogram(colour = "black", 
                 bins = 40, position="stack")+
 # facet_wrap("yearcol")+
  labs(title = "Distribution of estimated Input of Standing CWD", 
       fill="yearcol",
       y = "Count", 
       x = expression(paste("Log2 Mass (Mgha"^{-1},")"))) +  
  theme_bw() +  
  scale_y_continuous(breaks = seq(0, 20, by = 2))+
  scale_x_continuous(breaks = seq(-6, 10, by = 1))
  
  
  
```




**Figure 6**


Distribution of log₂-transformed input mass of standing CWD per subplot (excluding 2017), coloured by year. The x-axis represents log₂-transformed mass (Mg ha⁻¹), and the y-axis shows the count of subplots within each bin.  






### Variation in CWD stocks and fluxes over time within subplots
```{r, include=FALSE}
# Function to assign quantiles to the mass of fallen and standing CWD for each subplot
# Splitting into 10 groups based on maximum total mass for each subplot
QUANTILE<-function(indata) {
indata <- indata %>%
  dplyr::group_by(subplot_code) %>%
  dplyr::mutate(max_mass = max(mass.Mg.ha, na.rm = TRUE)) %>%
  ungroup()
indata <- indata %>%
  dplyr::mutate(mass_group = ntile(max_mass, 10)) %>%
  ungroup()
return(indata)
}
subplot_fallen<-QUANTILE(subplot_fallen)
subplot_standing<-QUANTILE(subplot_standing)

```




```{r, echo=FALSE, fig.width=10, fig.height=8}
# Plotting trends of fallen CWD mass for each subplot through time, split by mass quantile group
# Each subplot coloured uniquely
color <- grDevices::colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]
sampled_colors <- sample(color,100)

unique_subplots<-subplot_fallen$subplot_code
ggplot(subplot_fallen, aes (x=yearcol, y=mass.Mg.ha, colour = subplot_code))+
geom_point()+
geom_line()+
facet_wrap("mass_group",scales = "free", ncol=3)+
labs(title = "Temporal trajectories of mass of fallen CWD in subplots", 
      x = "Year", 
      y = expression(paste("Mass (Mgha"^{-1},")"))) +  
theme_bw() +
scale_x_continuous(breaks = seq(2017, 2024, by = 1))+
 # scale_y_continuous(breaks = seq( by = 2))+
theme(axis.text.x = element_text(),
      legend.position = "none",)+
scale_colour_manual(values = sampled_colors)

```



**Figure 7**
Interannual variation in estimated fallen CWD mass for each 40 × 40 m subplot, with subplots grouped by quantiles (specifically, deciles) of maximum total mass (for readability). Lines represent individual subplot trajectories over time. Note that the y axes differ for different groups of subplots.  





```{r, echo=FALSE, fig.width=10, fig.height=8 }
# Plotting trends of standing CWD mass for each subplot through time, split by mass quantile group
# Each subplot coloured uniquely

color <- grDevices::colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]
sampled_colors <- sample(color,100)

unique_subplots<-subplot_standing$subplot_code
ggplot(subplot_standing, aes (x=yearcol, y=mass.Mg.ha, colour = subplot_code))+
geom_point()+
geom_line()+
facet_wrap("mass_group",scales = "free", ncol=3)+
labs(title = "Mass of Standing CWD per Subplot Grouped by Maximum Total Mass Quantile", 
      x = "Year", 
      y = expression(paste("Mass (Mgha"^{-1},")"))) +  
theme_bw() +
scale_x_continuous(breaks = seq(2017, 2024, by = 1))+
 # scale_y_continuous(breaks = seq( by = 2))+
theme(axis.text.x = element_text(),
      legend.position = "none")+
scale_colour_manual(values = sampled_colors)
```




**Figure 8**

Annual variation in estimated standing CWD mass for each 40 × 40 m dynamic subplot, grouped by the subplot's maximum total mass quantile. Each subplot's trajectory is shown over time, coloured uniquely.



```{r, echo=FALSE}
# Function to categorize branchfall data and calculate mass and volume per year, subplot, and sample type
SAMPLETYPECALCULATIONS <- function(indata){
# Reassign branchfall data with descriptive labels
  # 0 = the piece is a branch
  # 1 = the piece is a trunk
  # 2 = the piece is unknown
  indata <- indata %>%
    dplyr::mutate(branchfall = dplyr::case_when(
      branchfall == 0 ~ "branch",
      branchfall == 1 ~ "trunk",
      branchfall == 2 ~ "unknown",
      TRUE ~ as.character(branchfall)
    ))

# Group data by year, subplot, and branchfall type, and calculate mass and volume
  indata <- indata %>%
    dplyr::group_by(yearcol, subplot_code, branchfall) %>%
    dplyr::summarise(
      mass.kg.m2 = (pi / (2 * 160)) * sum(pen_crossmass.kgm, na.rm = TRUE),
      vol.m3.m2 = (pi^2 / (8 * 160)) * sum(diam_sqr_m, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      mass.Mg.ha = mass.kg.m2 * 10,
      vol.m3.ha = vol.m3.m2 * 10000
    ) %>%
    dplyr::select(-mass.kg.m2, -vol.m3.m2)

  return(indata)
}

# Apply to your dataset
type_data <- SAMPLETYPECALCULATIONS(fallen_data)

# Function to add missing combinations of subplots, years, and branchfall types, filling missing values with zero
add_zeros <- function(indata, subplot, outdata) {
  all_subplots <- unique(subplot$subplot_code)
  all_years <- unique(indata$yearcol)
  all_branchfall <- c("unknown", "branch", "trunk")

  all_combinations <- expand.grid(subplot_code = all_subplots,
                                  yearcol = all_years,
                                  branchfall = all_branchfall,
                                  KEEP.OUT.ATTRS = FALSE)
  outdata <- merge(all_combinations, indata, 
                   by = c("subplot_code", "yearcol", "branchfall"), all.x = TRUE)
  outdata <- outdata %>%
    dplyr::mutate(across(where(is.numeric), ~replace(., is.na(.), 0)))
  return(outdata)
}


# Apply to data
type_data <- add_zeros(type_data, subplot, type_data)
# Calculate yearly average
mean_mass_by_year <- type_data %>%
  dplyr::group_by(yearcol, branchfall) %>%
  dplyr::summarise(mean_mass = mean(mass.Mg.ha, na.rm = TRUE), .groups = "drop")

# Calculating mean total mass of coarse woody debris by sample type (branch, trunk, unknown) for each year
# Using stacked bar plots to show contribution of each sample type
ggplot(mean_mass_by_year, aes(x =yearcol, y = mean_mass, fill = branchfall)) +
  geom_bar(stat = "identity", position = "stack", color="black") +
  labs(title = "Mean Total Mass of Coarse Woody Debris by sample type", 
      x = "Year", 
      y = expression(paste("Mass (Mgha"^{-1},")")),
      fill= "Sample Type") +
  scale_fill_manual(values=c( "skyblue","#882255", "#009E73" ))+
  scale_x_continuous(breaks = unique(mean_mass_by_year$yearcol))+ # Add titles  
  scale_y_continuous(limits=c(0,14), breaks = seq(0, 50, by = 1))+
  theme_bw()

```

**Figure 9**

Mean total mass of fallen CWD (Mg ha⁻¹) per year, grouped by sample type (branch, trunk, unknown). Bars represent the stacked mean mass of each component, illustrating changes in the composition of fallen woody debris through time.

rmarkdown::render("report_bci50CWD_2017-24.Rmd",
output_file = "Output/report_bci50CWD_2017-24.html")








